6.4132/16.413 Principles of Autonomy and Decision Making Final Project

### 2023-12-09 paper/main.tex
initial trajopt complete
yayayay im done

Activity planing, motion planning, and trajectory optimization for a robotic arm is challenging. To navigate around an environment, the agent must perform a state space search to plan a sequence of actions and generate a trajectory by either using a sample based planner or modeling the problem as a constraint optimization problem. This project explores 1) defining an activity planning problem with the Planning Domain Definition Language, 2) implementing the Fast-Forward heuristic planner with Enforced Hill Climbing, 3) using RRT with goal-biasing for trajectory generation, and 4) comparing the results with the trajectory generated by constraint optimization. The kitchen simulation environment provides an excellent sandbox for testing and demonstrating the effectiveness of the proposed approach, in addition to learning the benefits and drawbacks of the implemented algorithms.

@@

\iffalse
\subsection{Problem Statement}

The simulator consists of a kitchen environment with drawers, cabinets, countertops, and a robot manipulator arm. In addition, sugar and spam boxes are placed on a stovetop burner and a countertop. The objective is to move the sugar box to a nearby countertop and store the spam box inside the red drawer below it.

\begin{figure}[h]
    \centering
    \includegraphics[width=200px]{sim_env.jpg}
    \caption{The kitchen simulation environment.}
    \label{fig:sim_env}
\end{figure}
\fi

@@

The PDDL domain and problem files encode the problem as sets of objects, types, predicates, actions, and goals. There are two main types of objects: the \texttt{box} type and the \texttt{storage} type. The \texttt{box} type represents the sugar and spam boxes which are manipulated by the robot arm, and the \texttt{storage} type represents the storage spaces in the kitchen. In addition, I separated the \texttt{storage} type into two subtypes: \texttt{static}, which are storage spaces without actions (stovetop burner and countertop), and \texttt{openable}, which are storage spaces that can be opened or closed (drawer, cabinet).

@@

The \texttt{ff\_planner} file contains all functions to solve the activity planning problem. The \texttt{calculate\_hff} function computes the fast-forward heuristic $h_{ff}$ of a given state by counting the number of actions (without delete effects) it takes until the goal condition is a subset of \texttt{state}. At each iteration, \texttt{new\_state} is computed by taking the union of the current state and the add effects of all applicable actions. Because we defined our actions to only have positive preconditions, we can easily determine whether an action is applicable by checking if the precondition is a subset of \texttt{new\_state}. $h_{ff}$ is an admissible heuristic that always underestimates the number of actions it takes to reach the goal state, hence an alternative algorithm such as A* could have been used to find the optimal path.

In this project, I decided to implement Enforced Hill Climbing in the \texttt{solve\_ff} function. While EHC is not guaranteed to find the optimal solution, its greedy nature typically leads to higher efficiency and lower memory usage. \texttt{solve\_ff} initially parses the PDDL files and enumerates all possible actions that can be taken by the agent. Next, it performs a greedy search by immediately adding an action that leads to a strictly smaller $h_{ff}$ to the plan. If no such action is found, there is a ``plateau" in the graph, and BFS is used to find a sequence of actions that lead to a smaller $h_{ff}$. If BFS fails to find such action, then the algorithm determines that the activity planning problem is unsolvable.

Comparing the performance of BFS and FF Heuristic Planner, on the contrary to my expectation, I found that BFS consistently runs faster ($650\mu s$ vs. $950\mu s$). This is most likely because the plan graph is rather simple for this problem. For example, the starting state has $h_{ff}=2$, meaning EHC gets stuck on a ``plateau" for most of the time, resorting to BFS. In this case, calculating $h_{ff}$ leads to extra computation time. For more complex problems requiring many actions to be taken, I expect the FF Heuristic Planner to perform significantly faster than BFS.

@@

In addition, the following assumptions were made when implementing the motion planner:
\begin{itemize}
    \item The sugar/spam box attaches to the end-effector once the ``grab" action finishes
    \item The drawer opens after the ``open drawer" action ends and closes before the ``close drawer" action begins
\end{itemize}

@@

The sample-based motion planning algorithm is written in \texttt{rrt.py}. \texttt{get\_sample\_fn} returns a sampling function that returns a random valid joint configuration $x_\text{rand}$. I also defined a class \texttt{Tree} which keeps track of its \texttt{parent}, \texttt{children}, and value (\texttt{point}), which is the 7D joint configuration. \texttt{Tree} has function \texttt{add\_child}, which appends a new \texttt{Tree} to its \texttt{children} and sets the child's \texttt{parent} to itself.

Class \texttt{TrajectoryGenerator} contains the main motion planning algorithm in the \texttt{rrt} method. It also includes four helper functions:
\begin{enumerate}
    \item \texttt{find\_nearest}: returns $x_\text{nearest}$, the node in the tree closest to $x_\text{rand}$ by recursively traversing the tree.
    \item \texttt{steer}: creates a straight line from $x_\text{nearest}$ to $x_\text{rand}$ limited to length \texttt{JOINT\_STEP\_SIZE}. Returns this new point $x_\text{new}$.
    \item \texttt{obstacle\_free}: returns true if the input joint configuration does not cause the robot to collide with a kitchen object.
    \item \texttt{find\_path}: returns a list of joint configurations from the initial state to the goal state by following the parent pointers.
\end{enumerate}

@@

Instead of getting a random sample every time, the algorithm uses the goal state as the ``random" value every \texttt{GOAL\_SAMPLE} cycles to bias the tree towards the goal. Additionally, we determine whether $x_\text{new}$ is in the goal state by checking if the euclidian distance between $x_\text{new}$ and $x_\text{goal}$ is less than \texttt{GOAL\_THRESHOLD}.

@@

The \texttt{robot.py} file includes a \texttt{Robot} class with methods for each action in the PDDL domain. It also includes methods to move the robot base closer to the objects and the robot arm back to the initial state. Finally, the dictionary \texttt{function\_map} maps PDDL actions to its corresponding methods, and the \texttt{act} method uses this dictionary to perform a task based on the input PDDL action.

@@

@@

%%% TODO: EDIT REFERENCES %%%
\cite{Cui1}

>>>>

Activity planing, motion planning, and trajectory optimization for a robotic arm is challenging. To navigate around an environment, the agent must perform a state space search to plan a sequence of actions and generate a trajectory by using a sample-based planner or modeling the problem as an optimization problem. This project explores 1) defining an activity planning problem with the Planning Domain Definition Language, 2) implementing the Fast-Forward heuristic planner with Enforced Hill Climbing, 3) using RRT with goal-biasing for trajectory generation, and 4) comparing the results with the trajectory generated by constraint optimization. The kitchen simulation environment provides an excellent sandbox for testing and demonstrating the effectiveness of the proposed approach, in addition to learning the benefits and drawbacks of the implemented algorithms.

@@

@@

The PDDL domain and problem files encode the problem as sets of objects, types, predicates, actions, and goals. There are two main types of objects: the \texttt{box} type, which represents the sugar/spam boxes that are manipulated by the robot arm, and the \texttt{storage} type, which represents the storage spaces in the kitchen. In addition, the \texttt{storage} type is separated into two subtypes: \texttt{static}, which are storage spaces without actions (stovetop burner and countertop), and \texttt{openable}, which are storage spaces that can be opened or closed (drawer, cabinet).

@@

The \texttt{ff\_planner} file contains all functions to solve the activity planning problem. The \texttt{calculate\_hff} function computes the fast-forward heuristic $h_{ff}$ of a given state by counting the number of actions (without delete effects) it takes until the goal condition is a subset of \texttt{state}. At each iteration, \texttt{new\_state} is computed by taking the union of the current state and the add effects of all applicable actions. Because we defined our actions to only have positive preconditions, we can easily determine whether an action is applicable by checking if the precondition is a subset of \texttt{new\_state}.

In this project, I decided to implement Enforced Hill Climbing in the \texttt{solve\_ff} function. While EHC is not guaranteed to find an optimal solution, its greedy nature typically leads to higher efficiency and lower memory usage. \texttt{solve\_ff} initially parses the PDDL files and enumerates all possible actions that can be taken by the agent. Next, it performs a greedy search by adding the first action that leads to a strictly smaller $h_{ff}$ to the plan. If no such action is found, there is a ``plateau" in the graph, and BFS is used to find a sequence of actions that lead to a smaller $h_{ff}$. If BFS fails to find such sequence of actions, then the algorithm determines that the activity planning problem is unsolvable.

Comparing the performance of BFS and FF Heuristic Planner, on the contrary to my expectation, I found that BFS consistently runs faster ($650\mu s$ vs. $950\mu s$). This is because the plan graph is rather simple for this problem. For example, the starting state has $h_{ff}=2$, meaning EHC gets stuck on a ``plateau" for most of the time, resorting to BFS. In this case, calculating $h_{ff}$ leads to computational overhead. For more complex problems requiring many actions to be taken, I expect the FF Heuristic Planner to perform significantly faster than BFS.

@@

In addition, the following environment assumptions were made:
\begin{itemize}
    \item The sugar/spam box attaches to the end-effector once the ``grab" action finishes
    \item The drawer opens after the ``open drawer" action ends and closes before the ``close drawer" action begins
\end{itemize}

@@

The sample-based motion planning algorithm is written in \texttt{rrt.py}. \texttt{get\_sample\_fn} returns a sampling function that returns a random valid joint configuration $x_\text{rand}$. I also defined a class \texttt{Tree} which keeps track of its \texttt{parent}, \texttt{children}, and \texttt{point} (joint configuration), and has a method \texttt{add\_child} which appends a new \texttt{Tree} to its \texttt{children} and sets the child's \texttt{parent} to itself.

Class \texttt{TrajectoryGenerator} contains the main motion planning algorithm in the \texttt{solve} method. It also includes four helper functions:
\begin{enumerate}
    \item \texttt{find\_nearest}: returns $x_\text{nearest}$, the node in the tree closest to $x_\text{rand}$ by recursively traversing the tree.
    \item \texttt{steer}: creates a line from $x_\text{nearest}$ to $x_\text{rand}$ limited to length \texttt{JOINT\_STEP\_SIZE}. Returns this new point $x_\text{new}$.
    \item \texttt{obstacle\_free}: returns true if the input joint configuration does not cause the robot to collide with a kitchen object.
    \item \texttt{find\_path}: returns a list of joint configurations from the initial state to the goal state by following the parent pointers.
\end{enumerate}

@@

Instead of getting a random sample every time, the algorithm samples the goal state every \texttt{GOAL\_SAMPLE} cycles to bias the tree. Additionally, we determine whether $x_\text{new}$ is in the goal state by checking if the euclidian distance between $x_\text{new}$ and $x_\text{goal}$ is less than \texttt{GOAL\_THRESHOLD}.

@@

The \texttt{robot.py} file includes a \texttt{Robot} class with methods for each action in the PDDL domain in addition to methods to move the robot base and arm to the initial state. Dictionary \texttt{function\_map} maps PDDL actions to its corresponding methods, and the \texttt{act} method uses \texttt{function\_map} to perform a task based on a given action.

@@

To view the trajectory optimization in action, see \href{https://drive.google.com/file/d/1ZB1DUmpl6Mqf19IPNjAWVvW_RhoJtXB6/view?usp=sharing}{Trajectory optimization} or go to the following url: \url{https://drive.google.com/file/d/1ZB1DUmpl6Mqf19IPNjAWVvW_RhoJtXB6/view?usp=sharing}.

\subsection{File organization}

Trajectory optimization is implemented in pydrake using the default IPOPT nonlinear optimization solver, and can be found in \texttt{trajopt.py}. The \texttt{solve} function in the \texttt{TrajectoryOptimizer} class solves the optimization problem defined in \href{sec:opt}{Optimization problem definition} and takes an initial guess of the arm trajectory as an optional argument to seed the solver. Furthermore, the \texttt{Robot} class constructor has an additional \texttt{use\_trajopt} argument which allows the use of \texttt{TrajectoryOptimizer} to generate the arm trajectory. Finally, the trajectories generated from the sample-based motion planner were saved as \texttt{.npy} files in the \texttt{npy} folder. To use these trajectories as inputs to the IPOPT solver, I added a new \texttt{Robot().get\_init\_guess} method which loads a \texttt{.npy} file and linearly interpolates \texttt{PATH\_LENGTH} sample points. The solver generates a trajectory encoded as a $\mathtt{PATH\_LENGTH}\times7$ matrix, so the initial guess must also be of the same size.

\subsection{Optimization problem definition}
\label{sec:opt}

The kinematic trajectory optimization formulation is inspired from the problem used in Professor Russ Tedrake's Robotic Manipulation course \cite{Tedrake}:

\begin{equation*}
\begin{aligned}
    \min_T\quad & T,\\
    \textrm{subject to}\quad & X^{G_{start}}=f_{kin}(q(0)),\\
                             & X^{G_{goal}}=f_{kin}(q(T)),\\
                             & \forall t,\quad |\dot q|\le v_{max},
\end{aligned}
\end{equation*}

Where $f_{kin}$ is the forward kinematics function and $q(t)$ is the joint state of the robot at time $t$. The problem reads as minimizing time $T$ for the arm to move from the start pose $X^{G_{start}}$ to the goal pose $X^{G_{goal}}$ while satisfying velocity constraints.

There are a few challenges when converting this problem into a pydrake \texttt{MathematicalProgram}. Since pydrake does not allow pybullet functions to be included in its problem formulation, I simplified the problem to input initial and goal joint states rather than poses, and did not encode collision detection. Moreover, pydrake does not allow indexing into a variable using another variable. To avoid this problem, I hard-coded the length of the arm trajectory to be $\mathtt{PATH\_LENGTH}=200$ defined in \texttt{constants.py}. Finally, I approximated the joint velocities as the following: $\dot q\approx \frac{q(t+1)-q(t)}{T/\texttt{PATH\_LENGTH}}\quad\forall t\in\{0,\dots,T-1\}$. Bringing it all together, the new optimization problem is the following:

\begin{equation*}
\begin{aligned}
    \min_T\quad & T,\\
    \textrm{subject to}\quad & q_0=q[0],\\
                             & q_f=q[T],\\
                             & \forall t\in\{0,\dots,T-1\},\quad \frac{q[t+1]-q[t]}{T/\texttt{PATH\_LENGTH}}\le v_{max},\\
                             & \forall t\in\{0,\dots,T-1\},\quad \frac{q[t+1]-q[t]}{T/\texttt{PATH\_LENGTH}}\ge -v_{max}
\end{aligned}
\end{equation*}

where $q_0$ is the initial joint state and $q_f$ is the goal joint state. When running the program, the generated trajectories were smooth but inefficient as the solver found solutions with nonoptimal local minima. While this was alleviated by seeding the solver with initial guesses, it caused the optimized trajectories to be jagged since the given RRT trajectories exhibited a ``shaking" motion. To prevent such trajectories, I added another set of constraints to limit the acceleration of the joints:

\begin{equation*}
\begin{aligned}
    \forall t\in\{0,\dots,T-2\},\quad \frac{(q[t+2]-q[t+1])-(q[t+1]-q[t])}{2T/\texttt{PATH\_LENGTH}}\le a_{max},\\
    \forall t\in\{0,\dots,T-2\},\quad \frac{(q[t+2]-q[t+1])-(q[t+1]-q[t])}{2T/\texttt{PATH\_LENGTH}}\ge -a_{max}
\end{aligned}
\end{equation*}

Compared to the sample-based motion planner, the end result was smoother and more controlled trajectories but slower computation due to the greater number of constraints. Although some actions continued to cause the arm to move inefficiently, this is because the chioces of the goal joints can be improved. Implementing a custom forward kinematics function compatible with pydrake and setting the goal state as $X^{G_{goal}}=f_{kin}(q(T))$ will allow the program to find more optimal trajectories. Finally, to implement collision avoidance, I would encode the kitchen environment in pydrake and prevent the end-effector from colliding with the environment. However, I expect adding these two sets of constraints to significantly slow down computation.

@@

Using PDDL and the Fast-Forward heuristic planner is an efficient way to encode and solve an activity planning problem. Although the benefits of using the FF heuristic was unclear in this paper, it can improve performance when solving more complicated problems without many heuristic ``plateaus". Both the sample-based motion planner and nonlinear constraint optimization are valid ways to execute the generated plan. However, constraint optimization, with its ability to precisely encode robot dynamics, yielded a smoother and more optimized trajectory compared to the RRT implementation. Using complex algorithms such as RRT* may result in finding an optimal trajectory, but the required computation power may be too high. Similarly, adding more constraints to encode collision avoidance will result in slower computation for the nonlinear optimizer. For both activity and motion planning, it is essential to consider the compromise between runtime and finding a near-optimal solution, and choose the solution that benefits the program the most.
